ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-09-17 FinalReport/FinalReportDraft.tex
fixed while hanging (!!!!!) also have report in proper format from MA

\documentclass[12pt]{article}

\usepackage{natbib,hyperref,graphicx}
\usepackage[separate-uncertainty=true,multi-part-units=single]{siunitx}

\usepackage[margin=1.5cm, left=3cm ,twoside]{geometry}

@@

\title{ASTR480 Final Report:\\
\textit{Asteroids in TESS}
}
\author{Brayden Leicester \\ 
\textit{61352063} \\
[1ex]\small{Supervisors: Michele Bannister and Ryan Ridden-Harper}
}
\begin{document}
\pagenumbering{gobble} %to blank pg # before Contents pg
%TODO declaration page ????
%\newpage
\maketitle
%*{ as per the report guidelines
% \newpage
% \section*{Abstract}
% %*Background

@@

% \newpage
% \pagenumbering{roman}
% \tableofcontents
% \newpage
% \listoffigures
% \newpage
% \listoftables
% \newpage
%*}
\newpage
\pagenumbering{arabic}

@@

There have been attempts before to find and classify the asteroids in TESS data before.
\citet{Pal2018} proposes that TESS will be a good instrument for solar system object study, and simulate some detections down to \nth{20} visual magnitude, but that good photometry should only be expected to $V \lesssim 19$.
The first data release in \citet{Pal2020} catalogues nearly 10,000 asteroid light curves from the first year of TESS operation, and they report that they triple the number of asteroids with accurate rotation periods.
\citet{McNeill2023} also analyse TESS cycle 1, and detect almost 38,000 objects.
They determine reliable rotation periods for about 3,500 asteroids in this sample, and show lack of reliability for objects with periods less than \qty{3}{\hour}.
In the overlap from the data from \citet{Pal2020}, \citeauthor{McNeill2023} find good agreement between the two sets of periods and amplitudes.
\citet{Gowanlock2024} uses TESS photometry as well as  observations by the Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} to get a longer baseline on mutually observed objects while combining ground and space based observations.
There were not very many objects in both samples with known periods for comparison, with only 222 objects analysed, however they demonstrate that the technique is effective.

@@

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} %TODO updated in citation
in the Light Curve Database (LCDB).
%TODO fig of period vs diameter.   
In the October \nth{1} 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes data from \citet{Pal2020}.
Comparing the periods I find of asteroids that are already in the LCDB is key to making sure my methods are reliable, and that the periods for asteroids for which no data exists can be trusted.

@@

@@

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For most asteroids, their orbital elements are well known, so it is a matter of looking them up and cross-matching with transients in the TESS data.
Python was used to make API calls to {Skybot}\footnote{\href{https://vo.imcce.fr/webservices/skybot/}{Skybot: https://vo.imcce.fr/webservices/skybot/}} to get positions of asteroids in a cone search box in right ascension (RA) and  declination (Dec) space.
As TESS sectors are \qty{27}{\day} long, querying every \qty{12}{\hour} is manageable.
These positions are still very sparsely spaced in time compared to the TESS data, so an interpolation is needed to bridge the gap.

Knowing where the asteroids are is helpful for finding them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
  {JPL Horizons}\footnote{\href{https://ssd.jpl.nasa.gov/horizons/}{JPL Horizons: https://ssd.jpl.nasa.gov/horizons/}} was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$, which is a good proxy for size (as discussed above). %TODO ref

@@

\section*{Acknowledgements}
%OzSTAR asks for:
\small
This work was performed on the OzSTAR national facility at Swinburne University of Technology.
The OzSTAR program receives funding in part from the Astronomy National Collaborative Research Infrastructure Strategy (NCRIS) allocation provided by the Australian Government, and from the Victorian Higher Education State Investment Fund (VHESIF) provided by the Victorian Government.

%*Bib
\bibliographystyle{jphysicsB}
\bibliography{bibfile.bib}
\end{document}

>>>>

\documentclass{UCreport}

\usepackage[authoryear,comma,round]{natbib}
\usepackage{hyperref,graphicx}
\usepackage[separate-uncertainty=true,multi-part-units=single]{siunitx}

% \usepackage[margin=1.5cm, left=3cm ,twoside]{geometry}

@@

% \title{ASTR480 Final Report:\\
% \textit{Asteroids in TESS}
% }
% \author{Brayden Leicester \\ 
% \textit{61352063} \\
% [1ex]\small{Supervisors: Michele Bannister and Ryan Ridden-Harper}
% }
\begin{document}

%----------- Report information ---------

\logo{logos/UC.png}
\school{\textbf{School of Physical and Chemical Sciences}}
\course{\textbf{ASTR480}}
\student{\textbf{Brayden Leicester}}
\supervisor{Dr Michele Bannister and Dr Ryan Ridden-Harper}
\ttitle{Asteroids in TESS} %title of the file

%----------- Init -------------------

\buildmargins % display margins
\buildcover % create the front cover of the document

%------------ Declaration ----------------
\pagenumbering{roman}
\addcontentsline{toc}{section}{Declaration}
\declaration

I certify that content of this report was mostly my own work.
\\

My supervisor helped by proofreading the report and offering feedback. \\

Graduate students, X and Y, helped me understand ... \\

The derivation in section 3 was taken from ... \\

\vspace{2cm}
\begin{centering}
\textbf{\student}\par
\end{centering}

\newpage

%------------ Abstract ----------------

\addcontentsline{toc}{section}{Abstract}
\abstract

This is where the abstract goes. \\

This is the next paragraph.

% %*Background

@@

\newpage
%------------ Contents pages ----------------

\toc % creates the table of contents

\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\newpage
\addcontentsline{toc}{section}{List of Tables}
\listoftables
\newpage

%------------ Report body ----------------
\pagenumbering{arabic}

@@

There have been attempts to find and classify the asteroids in TESS data before.
\citet{Pal2018} proposes that TESS will be a good instrument for solar system object study, and simulate some detections down to \nth{20} visual magnitude, but that good photometry should only be expected to $V \lesssim 19$.
The sentiment of using TESS in this way is shared by \citet{Wong2019}.
The first data release in \citet{Pal2020} catalogues nearly 10,000 asteroid light curves from the first year of TESS operation, and they report that they triple the number of asteroids with accurate rotation periods.
\citet{McNeill2023} also analyse TESS cycle 1, and detect almost 38,000 objects.
They determine reliable rotation periods for about 3,500 asteroids in this sample and show a lack of reliability for objects with periods less than \qty{3}{\hour}.
In the overlap from the data from \citet{Pal2020}, \citeauthor{McNeill2023} find good agreement between the two sets of periods and amplitudes.
The Minor Planet Center (MPC) gets regular updates from TESS thanks to the work of \citet{Woods2021} and their \texttt{LINEAR-TESS} program, which creates tracklets of objects over a day, and then stitches these together to form a track of each asteroid through an FFI. 
\citet{Gowanlock2024} uses TESS photometry as well as  observations by the Zwicky Transient Facility \citep[ZTF, ][]{Bellm2018} to get a longer baseline on mutually observed objects while combining ground and space based observations.
There were not very many objects in both samples with known periods for comparison, with only 222 objects analysed, however they demonstrate that the technique is effective.
Fainter and unknown solar system objects can be found by shift and stacking \citep{Holman2019, Payne2019,Rice2020} or taking a fast X-ray transform \citep{Nguyen2024} of the FFI data.

@@

Because of such a rich history of study, many asteroids have a known rotation period.
These are catalogued by \citet{Warner2009} %TODO updated in citation http://www.MinorPlanet.info/php/lcdb.php
in the Light Curve Database (LCDB).
%TODO fig of period vs diameter.   
In the October \nth{1} 2023 release, the LCDB has nearly 9000 objects with periods determined to a quality code $U\geq 2-$, which is the minimum reliability recommended by \citeauthor{Warner2009} for statistical analysis.
This includes data from \citet{Pal2020}.
Comparing the periods I find of asteroids that are already in the LCDB is key to making sure my methods are reliable, and that the periods for asteroids for which no data exists can be trusted.

@@

% \begin{table}
%   \caption[Tab1]{My table}

% \end{table}

@@

To check for asteroids in the TESS data, the positions of the asteroids with time are required.
For most asteroids, their orbital elements are well known, so it is a matter of looking them up and cross-matching with transients in the TESS data.
Python was used to make API calls to {Skybot}\footnote{\href{https://vo.imcce.fr/webservices/skybot/}{Skybot: https://vo.imcce.fr/webservices/skybot/}} \citep{Berthier2006} to get positions of asteroids in a cone search box in right ascension (RA) and  declination (Dec) space.
As TESS sectors are \qty{27}{\day} long, querying every \qty{12}{\hour} is manageable.
These positions are still very sparsely spaced in time compared to the TESS data, so an interpolation is needed to bridge the gap.

Knowing where the asteroids are is helpful for finding them in the archival TESS data, but knowing more about the individual asteroids is also useful for population statistics.
The API of {JPL Horizons}\footnote{\href{https://ssd.jpl.nasa.gov/horizons/}{JPL Horizons: https://ssd.jpl.nasa.gov/horizons/}} was used to obtain the orbital elements of each asteroid, $a$, $e$, and $i$, as well as an absolute magnitude $H$, which is a good proxy for size (as discussed above). %TODO ref

@@

\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}
%OzSTAR asks for:
\small
This work was performed on the OzSTAR national facility at Swinburne University of Technology.
The OzSTAR program receives funding in part from the Astronomy National Collaborative Research Infrastructure Strategy (NCRIS) allocation provided by the Australian Government, and from the Victorian Higher Education State Investment Fund (VHESIF) provided by the Victorian Government.

This research has made use of data and/or services provided by the International Astronomical Union's Minor Planet Center.

This work has used data and/or services provided by NASA's Solar System Dynamics website, https://ssd.jpl.nasa.gov/. Specifically JPL Horizons: https://ssd.jpl.nasa.gov/horizons/

%*Bib
\addcontentsline{toc}{section}{References}
% \bibliographystyle{jphysicsB}
\bibliography{bibfile.bib}
\end{document}
